Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2017-01-21 draft.tex
.

\begin{document}
\maketitle
\section{Abstract} \label{abstract}
We attempted to replace the intrinsic homeostatic control system used in the original version by a mechanism based on the diffusion of a neurotransmitter across the nervous tissue based on the LIF-SORN-model proposed in \cite{SORN_Paper}. The model of diffusive homeostasis was adopted from a paper by Y. Sweeney et al. \cite{Sweeney_Paper} and models the tissue as a surface of square shape and a set of points on this surface, representing the neurons' positions within the SORN. The group of excitatory neurons then acted as a point-source of nitric oxide (NO), as well as as a sensor for the NO-concentration at each individual position. The production and sensing of NO forms the basis of a feedback loop: The individual NO-readout is fed into a comparator which causes an appropriate change within the internal firing threshold of the neuron, in turn altering the neuron's firing rate. The control system is then closed by linking the rate of NO-production to the neuron's firing rate.

Key aspects of this thesis include an analysis of the stability of the homeostatic control, followed by a comparison of features of the original LIF-SORN and the Diffusive variant. We expect to observe a preservation of non-random features that have been found in the original LIF-SORN while incorporating a stronger variance within neural activity (as reported by \cite{Sweeney_Paper}) which has previously been suppressed by a rigid single-cell homeostasis. In the face of possible \emph{new} features within the network's structure, one should further clarify the - presumably indirect - causal relation between diffusive spatial interaction and synaptic topology.

@@

Theoretical models of homeostasis are generally based on a form of feedback control. As such, it necessarily contains an error estimation. This could be based e.g. on the neuron's firing rate of the recent past measured in a given window, or, on a physiological level, e.g. a concentration of some molecule within some part of the cell structure. In any case, the error estimation requires setting some predefined target that the feedback control will attempt to achieve. This raises two questions: First, what would be an appropriate choice of homeostatic targets? Second, how could a neural system with fixed homeostatic targets encode changes of external stimuli if homeostasis overrides any information about changes in synaptic input?

@@

Since our main goal of the implementation of diffusive homeostasis was to allow the network to develop a broader distribution of firing rates across excitatory neurons compared to the original version of homeostasis, we first present the results of a comparison between features of the network activity in both variants in section \ref{activ_analys}. Following this is an analytic discussion of an instability we observed within the diffusive homeostatic feedback loop, see section \ref{theor_osc}. Furthermore, we compare topological features of the network under the influence of diffusive and non-diffusive homeostasis in section ???.

@@

\subsubsection{Distribution of Firing Rates}\label{Fir_Dist_Section}
Achieving a broad distribution of firing rates among neurons was the core motivation for the implementation of diffusive homeostasis. Figure \ref{Fir_Rate_Dist_Compare} shows a first result, comparing both homeostatic mechanisms. As expected, non-diffusive homeostasis leads to a sharp distribution of firing rates at 3 Hz. Diffusive homeostasis indeed results in a much broader distribution of mean firing rates. A large number of experimental studies have found that distributions of firing rates are not only broadly distributed but well described by a log-normal distribution, which has a non-zero third moment or skewness (\cite{Buzsaki_Fir_Rates_2014}, \cite{Wohrer_Fir_Rates_2012}). By definition, the logarithm of the random variable in question is thus again normally distributed. To check for this property we plotted the the distribution of decadic logarithms of firing rates in Figure \ref{Fir_Rate_Dist_Compare} (B). In (A), we found a skewness of $\mathrm{v_{Diff} = 0.765}$, in (B) $\mathrm{v_{Diff,log} = -0.488}$. Though this told us that the distribution is "more symmetric" on a logarithmic scale, it should rather be seen as being neither strictly normally or log-normally distributed.  
\begin{figure}
%%% plot_frequ_dist_mult_sets.py
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/fir_rate_dist_e_compare.png}
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/fir_rate_dist_i_compare.png}
\caption{Histograms of mean firing rates over the excitatory/inhibitory population in regular (A/A*) and logarithmic space (B/B*). For diffusive homeostasis ($\mathrm{D=1000\, \mu m^2 s^{-1}}$), the distribution was generated from 10 simulation runs, 1 simulation was used for non-diffusive homeostasis. Mean firing rates were calculated from spikes within $\mathrm{t=1000-1500\,s}$.}
\label{Fir_Rate_Dist_Compare}
\end{figure}
In contrast, inhibitory cells showed a much clearer distinction: A skewness of $\mathrm{v_{Diff} = 1.522}$ in regular space and $\mathrm{v_{Diff,log} = -0.147}$ in log-space. As a first interpretation, we related this difference to the fact that inhibitory cells are only indirectly affected by diffusive homeostasis in the sense that their distribution of activity is mainly determined by differences withing neural input while inhibitory thresholds are all initialized and kept at the same value.

@@

\subsection{Network Topology}
So far, we have only analyzed network activity, whereas in this section we present results that relate to features of synaptic topology. Since the SORN we used for all simulations only included plastic mechanisms affecting recurrent excitatory connections, these were the main subject of our investigations. Moreover, in contrast to the previous section, we did not aim for the emergence of an explicitly \emph{new} feature of synaptic topology, but rather sought to recover properties that had been found in earlier versions of the LIF-SORN.

>>>>

\begin{document}
\maketitle

@@

Theoretical models of homeostasis are generally based on a form of feedback control. As such, it necessarily contains an error estimation. This could be based e.g. on the neuron's firing rate of the recent past measured in a given window, or, on a physiological level, e.g. a concentration of some molecule within some part of the cell structure. In any case, the error estimation requires setting some predefined target that the feedback control will attempt to achieve. This raises two questions: First, what would be an appropriate choice of homeostatic targets? Second, given this set of targets, how could a neural system with fixed homeostatic targets encode changes of external stimuli if homeostasis overrides any information about changes in synaptic input?

A possible way of choosing homeostatic targets could be to randomly draw from a distribution that matches statistics known to be found in the neural tissue that one wishes to model. This, however is a highly artificial ad-hoc solution. Furthermore, it does not resolve the second issue of encoding of external input.

This thesis is based on earlier versions of a self-organizing recurrent neural network (SORN), in particular an implementation using leaky integrate-and-fire neurons (LIF-SORN) \cite{SORN_Paper}. It was developed to model features of activity and topology of cortical networks, in particular L5 of the neocortex. In these previous implementations, a slow homeostatic mechanism was implemented among excitatory neurons by an intrinsic adaptation of excitability. This led to the aforementioned problems of homeostatic control: Neural activity was tuned to a globally set target, allowing for no variability. In contrast, cortical networks are known to exhibit a broad distribution of firing rates, which is well described by a right-skewed, log-normal like shape \cite{Buzsaki_Fir_Rates_2014,Wohrer_Fir_Rates_2012}. If the LIF-SORN was meant to provide a good model for cortical structure and activity, it would require a different type of homeostatic model.

In this thesis, we examined \textit{diffusive homeostasis} as a possible candidate to replace the previously used model of intrinsic plasticity. The idea and modelling of diffusive homeostasis was adopted from a paper by Y. Sweeney et al. \cite{Sweeney_Paper}, which models neural tissue as a two-dimensional surface and a set of points representing the neurons' positions. The group of excitatory neurons acted as a point-source of nitric oxide (NO), as well as as a sensor for the NO-concentration at each individual position. The production and sensing of NO forms the basis of a feedback loop: The individual NO-readout is fed into a comparator which causes an appropriate change within the internal firing threshold of the neuron, in turn altering the neuron's firing rate. The control system is then closed by linking the rate of NO-production to the neuron's firing rate. The results by Sweeney et al. suggest that diffusive signalling could resolve the dichotomy between overall stability of firing activity and the need to allow flexibility among individual neurons' activities. Intuitively speaking, through the diffusive signal, each neurons receives a "mixture" of its own activity and its neighbouring neurons. Thereby individual tuning of firing rates is suppressed while the overall population activity is kept at a constant level.

Key aspects of this thesis include an analysis of the stability of the homeostatic control, followed by a comparison of features of the original LIF-SORN and the diffusive variant. We expect to observe a preservation of non-random features that have been found in the original LIF-SORN while incorporating a stronger variance within neural activity which has previously been suppressed by the rigid single-cell homeostasis. In the face of possible \emph{new} features within the network's structure, one should further clarify the causal relation between diffusive spatial interaction and synaptic topology.

@@

Since our main goal of the implementation of diffusive homeostasis was to allow the network to develop a broader distribution of firing rates across excitatory neurons compared to the original version of homeostasis, we first present the results of a comparison between features of the network activity in both variants in section \ref{activ_analys}. Following this is an analytic discussion of an instability we observed within the diffusive homeostatic feedback loop, see section \ref{theor_osc}. Furthermore, we compare topological features of the network under the influence of diffusive and non-diffusive homeostasis in section \ref{Network_Topology_Section}.

@@

\subsubsection{Distribution of Firing Rates}\label{Fir_Dist_Section}
Achieving a broad distribution of firing rates among neurons was the core motivation for the implementation of diffusive homeostasis. Figure \ref{Fir_Rate_Dist_Compare} shows a first result, comparing both homeostatic mechanisms. As expected, non-diffusive homeostasis leads to a sharp distribution of firing rates at 3 Hz. Diffusive homeostasis indeed results in a much broader distribution of mean firing rates. As mentioned in the introduction, a large number of experimental studies have found that distributions of firing rates are not only broadly distributed but well described by a log-normal distribution, which has a non-zero third moment or skewness. By definition, the logarithm of the random variable in question is thus again normally distributed. To check for this property we plotted the the distribution of decadic logarithms of firing rates in Figure \ref{Fir_Rate_Dist_Compare} (B). In (A), we found a skewness of $\mathrm{v_{Diff} = 0.765}$, in (B) $\mathrm{v_{Diff,log} = -0.488}$. Though this told us that the distribution is "more symmetric" on a logarithmic scale, it should rather be seen as being neither strictly normally or log-normally distributed.  
\begin{figure}
%%% plot_frequ_dist_mult_sets.py
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/fir_rate_dist_e_compare.png}
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/fir_rate_dist_i_compare.png}
\caption{Histograms of mean firing rates over the excitatory/inhibitory population in regular (A/A*) and logarithmic space (B/B*). For diffusive homeostasis ($\mathrm{D=1000\, \mu m^2 s^{-1}}$), the distribution was generated from 10 simulation runs, 1 simulation was used for non-diffusive homeostasis. Mean firing rates were calculated from spikes within $\mathrm{t=1000-1500\,s}$.}
\label{Fir_Rate_Dist_Compare}
\end{figure}
In contrast, inhibitory cells showed a much clearer distinction: A skewness of $\mathrm{v_{Diff} = 1.522}$ in regular space and $\mathrm{v_{Diff,log} = -0.147}$ in log-space. As a first interpretation, we related this difference to the fact that inhibitory cells are only indirectly affected by diffusive homeostasis in the sense that their distribution of activity is mainly determined by differences withing neural input while inhibitory thresholds are all initialized and kept at the same value.

@@

\subsection{Network Topology}\label{Network_Topology_Section}
So far, we have only analyzed network activity, whereas in this section we present results that relate to features of synaptic topology. Since the SORN we used for all simulations only included plastic mechanisms affecting recurrent excitatory connections, these were the main subject of our investigations. Moreover, in contrast to the previous section, we did not aim for the emergence of an explicitly \emph{new} feature of synaptic topology, but rather sought to recover properties that had been found in earlier versions of the LIF-SORN.
